---
title: "LFS Variable Codebook (5q longitudinal)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LFS Variable Codebook (5q longitudinal)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(lfsclean)
```

## List of Variables

This document details the cleaned variables derived from the five-quarter longitudinal Labour Force Survey datasets using the `lfsclean` package. The longitudinal datasets contains a subset of the variables from the cross-sectional dataset and takes a single cohort of respondents to the LFS, restricted to the respondents who responded in each of the five quarters of their participation in the survey. 

The data are by default cleaned and output by the `lfsclean` functions in "wide" form, which means a single row for each individual and a column for each variable in each wave. 

## Time-invariant Variables

Variables which are constant across the waves, and therefore there is only one entry.

- __persid__ (numeric). Individual identifier.

- __lgwt22__ (numeric). Longitudinal weight for producing population-representative statistics.

- __year__ (numeric). Year of first wave interviewed.

- __quarter__ (numeric). Calendar quarter of first wave interviewed.

- __sex__  (factor):
  - "male"
  - "female"

- __empl_sequence__ (character). 5-character representation of the respondents labour market status (ILO definition) over the five waves.e.g. "EEEEU" is an individual employed for the first four quarters and then unemployed in the fifth quarter.

  - "E" employed / self-employed
  - "U" ILO unemployed
  - "I" Inactive
  


## Time-varying Variables 

Variables whose values can change between waves.

### Demographic 

- __age(1-5)__ (numeric). Age in single years 

- __eth4cat(1-5)__ (factor). Ethnicity (4 categories)
  - "white"
  - "black"
  - "mixed"
  - "asian_other"

- __empstat2cat(1-5)__ (factor). Economic status (2 categories)
  - "employed" (including self-employed)
  - "not_employed"

- __empstat3cat(1-5)__ (factor). Economic status (3 categories)
  - "employed" (including self-employed)
  - "inactive" (including those in education)
  - "unemployed"

- __empstat8cat(1-5)__ (factor). Economic status (8 categories)
  - "employed"
  - "self_employed"
  - "unemployed"
  - "retired"
  - "education"
  - "caring"
  - "sick"
  - "other"

- __region(1-5)__
  - "north_east"
  - "north_west"
  - "yorks_and_humber"
  - "east_mids"
  - "west_mids"
  - "east_of_england"
  - "london"
  - "south_east"
  - "south_west"
  - "scotland"
  - "wales"
  - "northern_ireland"
  
### Hours and earnings 

- __grsswk(1,5)__ (numeric). Gross nominal weekly earnings of the employed/self-employed (note, only available in waves 1 and 5. Not asked in waves 2-4)

- __uhours(1-5)__ (numeric). Hours usually worked per week (employed)

### Benefits and sickness

- __benclaim(1-5)__ (factor). Reason for individual receiving benefits. 
  - "jobseeker"
  - "sick"
  - "lone_parent"
  - "carer"
  - "other_oow_benefits"
  - "other_benefits"
  - "no_benefits"
  

- __disab(1-5)__ (factor). Disabled under the Equality Act. 
  - "disabled"
  - "not_disabled"

- __numsickdays(1-5)__ (numeric). The number of sick days the individual person has taken off in the last week (if any)
