---
title: "LFS Variable Codebook (cross-sectional)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LFS Variable Codebook (cross-sectional)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(lfsclean)
```

## List of Variables

This document details the cleaned variables derived from the cross-sectional quarterly Labour Force Survey datasets using the `lfsclean` package, grouped by cleaning function.

## Demographic variables

Demographic variables produced by the `lfs_clean_demographic` function.

- __d_age__ (numeric)

- __d_sex__  (factor):
  - "male"
  - "female"

- __d_ethnicity_4cat__  (factor):
   - "white"
   - "mixed"
   - "black"
   - "asian_other"

- __d_ethnicity_2cat__  (factor):
  - "white"
  - "non_white"

- __d_region__  (factor):
  - "north_east"
  - "north_west"
  - "yorks_and_humber"
  - "east_midlands"
  - "west_midlands"
  - "east_of_england"
  - "london"
  - "south_east"
  - "south_west"
  - "wales"
  - "scotland"
  - "northern_ireland"
  
- __d_country__ (factor):
  - "england"
  - "scotland"
  - "wales"
  - "northern_ireland"

- __d_marstat__ (factor):
  - "single"
  - "married"
  - "sep_div_wid"


## Economic activity 

Variables cleaned by the `lfs_clean_economic_activity` function - economic status and job characteristics for those in work.

- __l_lmstatus_2cat__ (factor):
  - "employed"
  - "non_employed"

- __l_lmstatus_3cat__ (factor):
  - "employed"
  - "unemployed"
  - "inactive"
  
- __l_lmstatus_8cat__ (factor):
  - "employed"
  - "self_employed"
  - "retired"  
  - "unemployed"
  - "sick"
  - "caring"
  - "education"
  - "other"
  
## Education 

Qualification variables cleaned by the `lfs_clean_education` function.

- __q_highest_qual__ (factor):
  - "Degree"
  - "Level 4+ Vocational"
  - "Level 3"
  - "Level 2"
  - "Below Level 2/Other"
  - "No qualifications"
  
- __q_edage__ (numeric):

## Earnings and hours 

For those in work, variables on hours worked (and would like to work) and earnings cleaned by the `lfs_clean_hours_earn` function.

- __eh_awage__ Real hourly earnings in March 2020 prices, based on actual hours worked in the reference week (Numeric). 

- __eh_uwage__ Real hourly earnings in March 2020 prices, based on usual hours worked (Numeric). 

- __eh_weekly_earnings__ Gross real weekly earnings in March 2020 prices in the reference week (Numeric).

- __eh_awage_nom__ Nominal hourly earnings, based on actual hours worked in the reference week (Numeric).

- __eh_uwage_nom__ Nominal hourly earnings, based on usual hours worked (Numeric).

- __eh_weekly_earnings_nom__ Gross nominal weekly earnings in the reference week (Numeric).

- __eh_ahours__ Hours worked in the reference week (Numeric).

- __eh_uhours__ Hours usually worked in a week (Numeric).

- __eh_undemp__ Would like to work more hours? (Factor)
  - "yes"
  - "no"

- __eh_ovemp__ Would like to work fewer hours? (Factor)
  - "yes"
  - "no"
  
- __eh_ovhrs__ Number of horus fewer would like to work (Numeric)

- __eh_undhrs__ Number of hours more would like to work (Numeric)
