---
title: "LFS Variable Codebook (5q longitudinal)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LFS Variable Codebook (5q longitudinal)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(lfsclean)
```

## List of Variables

This document details the cleaned variables derived from the five-quarter longitudinal Labour Force Survey datasets using the `lfsclean` package. The longitudinal datasets contains a subset of the variables from the cross-sectional dataset and takes a single cohort of respondents to the LFS, restricted to the respondents who responded in each of the five quarters of their participation in the survey. 

The data are by default cleaned and output by the `lfsclean` functions in "wide" form, which means a single row for each individual and a column for each variable in each wave. 

## Time-invariant Variables

Variables which are constant across the waves, and therefore there is only one entry.

- __persid__ (numeric). Individual identifier.

- __lgwt22__ (numeric). Longitudinal weight for producing population-representative statistics.

- __sex__  (factor):
  - "male"
  - "female"

## Time-varying Variables 

Variables whose values can change between waves.

- __age(1-5)__ (Numeric). Age in single years 

- __etukeul(1-5)__ (Factor). Ethnicity

- __ilodefr(1-5)__ (Factor). Economic status (ILO definition)


